**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-08-29 submission/manuscript.Rmd
mini review

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3]. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models. These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism. Furthermore, it was encouraging that even for a small number of samples, the 95% confidence interval included the median AUROC values for the larger subsetted datasets.

@@

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed at the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having SRN. Among the highest ranked OTUs across these three models, it was encouraging that many of the same OTUs (e.g. OTU 50, 426, 609, 822, 1239) were important across all of the models. The benefit of this approach is that the results of the analysis are based on the trained model parameters and provide information regarding the sign and magnitude of the impact of each OTU. However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact for each of these models, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 were permuted across the diagnoses from the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if we permuted the same OTU from the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`. Effectively, the complexity of the communities was more fully represented in the better performing models[@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. While permutation analysis allowed us to gauge the importance of an OTU, the analysis is post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.

@@

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. As expected, the training times increased with the complexity of the model and the number of tuned hyperparameter settings. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5]. When we subsetted the size of the training dataset, we observed an exponential relationship between the size of the dataset and the training time for the random forest model [Figure S5].

@@

Overview
•	Establish pipeline that can be generalized to any modeling method
•	Linear/non-linear not a bfd
•	Interpretability tradeoff
Point 1
•	Importance for transparency, picking the most complex model is not necessarilty the best model
•	Good / better / best:  Table S1
•	“best” has multiple meanings
Point 2
•	Our paper is concerned with developing models to gain greater biological insights into what’s going on
•	Still need biological validation and further experimentation

Point 3
•	These models show great opportunity to use microbiome to make diagnoses
•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
•	We also have a problem with OTU-based approaches
Conclusion/Future direction
•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out. We encourage reseraches to apply multiple models to their data and report the results
•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations. Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification. Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models where we pointed out potential pitfalls and ways to avoid them [Table 2].

To test this framework, we benchmarked seven ML models to show that a clearly defined ML problem that is based on the goal of the microbiome study should inform our model selection. Our results showed that if the goal of a study to identify microbial biomarkers of disease, we can create ML models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a more complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights.

Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease. Human-associated microbial communities have complex correlation structures. This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013]. In this study we used two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. The feature weights of correlated OTUs are influenced by one another. For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight in the regression model. To avoid misinterpreting the models and missing important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we avoided misinterpreting models due to correlation structures of the communities by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.

We established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We performed an initial 80%-20% random datasplit to create an held-out test set which we used to evaluate model performance. We repeated this datasplit 100 times to get a robust reading of predictive performance. We tuned our model hyperparameters in a five-fold cross-validation. The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.

We used a CRC dataset to develop ML models with a binary classification task. We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes. However, the principles highlighted throughout this study [Table 2] apply to all ML modeling tasks with microbiome data. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting.

>>>>

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3]. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models. These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism. Furthermore, it was encouraging that even for a small number of samples, the interquartile range included the median AUROC values for the larger subsetted datasets.

@@

First, we interpreted the feature importance of the linear models (L1 and L2-regularized SVM with linear kernel and L2-regularized logistic regression) using the median rank of absolute feature weights for each OTU [Figure 3]. We also reviewed the signs of feature weights to determine whether an OTU is associated with classifying a subject as being healthy or having an SRN - negative sign indicated being healthy and positive sign indicated having an SRN. It was encouraging that many of the highest ranked OTUs were shared across these three models, (e.g. OTU 50, 426, 609, 822, 1239). The benefit of this approach was that the results of the analysis were based on the trained model parameters and provided information regarding the sign and magnitude of the impact of each OTU. However, this approach was only possible with linear models.

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact for each of these models, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 were permuted across the diagnoses from the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if we permuted the same OTU from the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(round(RF$imp, 3))`. Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. While permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU is associated with classifying a subject as being healthy or having an SRN.

@@

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. As expected, the training times increased with the complexity of the model and the number of tuned hyperparameter settings. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5]. When we subsetted the size of the training dataset, we observed a linear relationship between the size of the dataset and the training times for L2-regularized logistic regression and random forest models [Figure S5].

@@

Overview
•	Establish pipeline that can be generalized to any modeling method
•	Linear/non-linear not a bfd
•	Interpretability tradeoff

Point 3
•	These models show great opportunity to use microbiome to make diagnoses
•	We talked about classification, but others have used it in microbiome data to do regression, mulit-class – pipeline is the same
•	Others have used this on shotgun metagenomic data and collinearity is expected to be an even larger problem since 5000 genes are all correlated with each other
•	Deployment is a different question that requires independent datasets to test on; we currently lack decent sized parallel datasets where the same type of sequence data and clinical data have been collected
•	We also have a problem with OTU-based approaches

Conclusion/Future direction
•	ML is awesome and there’s always new methods; our framework gives structure to investigators wanting to try these out. We encourage reseraches to apply multiple models to their data and report the results
•	Things like neural nets are attractive and hold great promise, but we need a ton more data to apply and they are impossible to interpret

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations. Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification. Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models and we highlighted potential pitfalls and ways to avoid them [Table S1].

The foundation of this framework was to clearly define the ML task and to choose the best model for that task [Table S1]. We benchmarked seven ML models and showed that if the ML task is to identify microbial biomarkers of disease, we can create models that are inherently interpretable and easily trained without losing predictive power. In terms of predictive performance, random forest model had the best testing AUROC values compared to the other six models. However, the second best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest was a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model is not necessarilty the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome [Table S1]. We performed an initial 80%-20% random datasplit to create an held-out test set which we used to evaluate model performance. We repeated this datasplit 100 times to get a robust reading of predictive performance. We tuned our model hyperparameters in a five-fold cross-validation. The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures. This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013]. The feature weights of correlated OTUs are influenced by one another. For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight thus hiding the importance of its correlated OTU. To avoid misinterpreting the models and to capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we avoided misinterpreting models by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features. To identify the true underlying microbial factors of a disease, it is crucial to do further experimentation for biological validation.

We did not evaluate multicategory classification methods or regression analyses to predict non-binary outcomes. However, the principles highlighted throughout this study [Table S1] apply to all ML modeling tasks with microbiome data. Our analysis was also limited to shallow learning methods and did not explore deep learning methods such as neural networks. Microbiome datasets often suffer from having high dimensionality but low sample sizes which makes deep learning models prone to overfitting.
